Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-16 talk/dls2012/paper.tex

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimizations
are simple requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically perform quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.
In this paper we explain a scheme invented within the context of the LuaJIT project
for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves.
We have implemented the scheme in PyPy's tracing JIT compiler,
where it can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
attributes.\footnote{This is sometimes called \emph{scalar replacement}~\cite{kotzmann_escape_2005}.}
If some of the attributes are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed from $\hat J$ using the map $m$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to
the trace in Figure~\ref{fig:virtual-trace}.

@@

\subsection{Python}
The Python interpreter of the RPython framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python, C and Lua and their
runtimes are compared in Figure~\ref{fig:benchmarks}. The benchmarks are

@@

  there are three different implementations.

@@

is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_n\right)$ from a vector

@@

$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n$. Both the output vector, $\bf b$, 

@@

for $1 \leq i \leq m$ and $1 \leq j \leq n$.

@@

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a straightforward implementation providing 2 dimensional
indexing with out of bounds checks and 
data stored in row-major order.
For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C. All the benchmarks except sqrt operate on C double-precision floating
point numbers, both in the Python and the C code.

@@

For PyPy 10 iterations were run, prefaced with 3 iterations for warming up.
Due to benchmarks taking large amounts of time on CPython, only one run
was performed, prefaced with one warmup run for Psyco.
For GCC 5 iterations
were run. In all cases, the standard deviation is very low, making benchmarks
very well reproducible.

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of RPython's JIT machine code backend as well as missing
optimizations, like instruction scheduling.

@@

The type specialization described by Gal \etal~\cite{gal_trace-based_2009} can
be seen as doing a similar optimization (again by manually implementing it)
than the one described in Section~\ref{sub:allocation}: The effect of both is
that type checks are fully done before a loop is even entered.

>>>>

\begin{abstract}
One of the nice properties of a tracing JIT is that many of its optimizations
are simple, requiring one forward pass only. This is not true for loop-invariant code
motion which is a very important optimization for code with tight kernels.
Especially for dynamic languages that typically perform quite a lot of loop invariant
type checking, boxed value unwrapping and virtual method lookups.
In this paper we explain a scheme pioneered within the context of the LuaJIT project
for making simple optimizations loop-aware by
using a simple pre-processing step on the trace and not changing the
optimizations themselves.
We have implemented the scheme in PyPy's tracing JIT compiler,
where it can give performance improvements of a
factor over two for PyPy's Python JIT executing simple numerical kernels
bringing the performance close to that of compiled C code.
\end{abstract}

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
attributes.\footnote{This is sometimes called \emph{scalar replacement}~\cite{kotzmann_escape_2005}.}
If some of the attributes are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed from $\hat J$ using the map $m$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $J$ is replaced by $\hat J$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to
the trace in Figure~\ref{fig:virtual-trace}.

@@

\subsection{Python}
The Python interpreter of the RPython framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python, C and Lua and their
runtimes are compared in Figure~\ref{fig:benchmarks}.\footnote{
    The benchmarks and the scripts to run them can be found in the repository for this paper:
    \texttt{https://bitbucket.org/pypy/extradoc/src/ tip/talk/dls2012/benchmarks}
}

The benchmarks are

@@

  there are three different implementations. In Lua there is no support for
  integers so only two versions are provided: float and Fix16. Here Fix16 is a custom class
  that implements scaled floating point arithmetic.

@@

is used to calculate a vector ${\bf b} = \left(b_1, \cdots, b_{n-2}\right)$ from a vector

@@

$b_i = k_3 a_i + k_2 a_{i+1} + k_1 a_{i+2}$ for $1 \leq i \leq n-2$. Both the output vector, $\bf b$, 

@@

for $2 \leq i \leq m-1$ and $2 \leq j \leq n-1$.

@@

The sobel and conv3x3 benchmarks are implemented
on top of a custom two-dimensional array class.
It is
a straightforward implementation providing 2 dimensional
indexing with out of bounds checks and 
data stored in row-major order.
For the C implementations it is
implemented as a C++ class. The other benchmarks are implemented in
plain C. All the benchmarks except sqrt operate on C double-precision floating
point numbers, both in the Python, C and Lua code.

@@

For PyPy and Lua 10 iterations were run, prefaced with 3 iterations for warming up.
Due to benchmarks taking large amounts of time on CPython, only one run
was performed, prefaced with one warmup run for Psyco.
For GCC 5 iterations
were run. In all cases, the standard deviation is very low, making benchmarks
very well reproducible.

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of RPython's JIT machine code backend as well as missing
optimizations, like instruction scheduling. Also, in case of nested loops, 
operations are only moved out of the 
innermost loop. That is an issue when the innermost loop is 
short and a significant amount of time is spent in the outer loops. This is the case 
with for example SparseMatMult.

@@

The type specialization described by Gal \etal~\cite{gal_trace-based_2009} can
be seen as doing a similar optimization (again by manually implementing it)
as the one described in Section~\ref{sub:allocation}: The effect of both is
that type checks are fully done before a loop is even entered.
